Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-04-03 paper1_jiturra.Rmd
update paper

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
# date: "April 2024"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | The escalating economic inequality and the current sanitary crisis have threatened social solidarity among citizens. However, these consequences have been unevenly experienced across the social structure where the upper and intermediate classes, benefiting from privileged access to social resources, have shown resilience. Conversely, working-class families have faced deteriorating material conditions, leading to a heightened sense of marginalization and increasing the demand for welfare support. Particularly, social resources through network ties have been suggested to have a direct link with an individual's welfare through instrumental and expressive outcomes, such as providing information or help in moments of need. Consequently, it has been argued that social isolation not only plays a role in terms of social resources, but higher levels of segregation can bolster opinions, suggesting that being isolated from other social classes can polarise attitudes mainly in the working and upper-middle classes. In this regard, based on the ISSP 2017 -  Social Networks (N=31,191),  multilevel estimations suggest that the influence of network homogeneity is conditional to social class, where the working and intermediate classes with high homogeneous networks hold higher redistributive preferences than the upper classes. Additionally, income inequality has been shown to decrease the polarizing effect of network segregation, mainly among the upper classes, when overall inequality is high. In sum, this study highlights the role of both social class and network segregation together to understand redistributive preferences, as well as their implications for policies and cohesion.
  | 
  | **Keywords**: network segregation, redistributive preferences, economic inequality
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

Individual data from the International Social Survey Programme (ISSP) 2017, titled "Social Networks and Social Resources," is utilized [@issp_research_group_international_2019]. Established in 1984, the ISSP is a collaborative research project that conducts cross-national surveys on various topics related to social issues, attitudes, and values. Specifically, the ISSP 2017 survey includes questions on social ties and activities, evaluated through a position generator, as well as inquiries on social resources derived from network members. Additionally, respondents express their views on topics such as social trust and attitudes toward economic inequality. Initially, the dataset comprised 47,027 individual observations across 32 countries. However, Slovenia (SVN) was excluded from the analysis because the key dependent variable was not present in the questionnaire. Consequently, the present study comprises complete information on 31,191 individuals from 31 countries, representing a diverse range of economic, cultural, and institutional characteristics.

@@

\pagebreak 
## Income Inequality Ratios

@@

@@

# Social distance

**Social distance**: First, the average ISEI points for each occupation of the position generator are calculated. Second, the ISEI of the respondent (R's) is subtracted from the average ISEI points of the personal network. For example, if the R's has an ISEI of 80 and the network ISEI is 50, the social distance will be 30 (80 - 50), a "upward" social distance. Another case could be 50 (R's) minus 80 (network), and the average social distance will be - 30 or "downward" social distance. In addition, When the distance is 0, the network is entirely homogeneous.

@@

## Social Distance by Social Class

@@

@@

@@

```{r fig1-marg-socdistclass, echo=FALSE, fig.cap='Average Marginal Effects of Social Distance conditionated by Social Class on Redistributive Preferences', fig.dim=c(10,5), warning=FALSE}
int_homo <- digclas1_VW_7
plot_slopes(digclas1_VW_7, variables = "homclass3_V_res", condition = list("dclass3res_V"))+
labs(caption = paste0("N=",nobs(int_homo),", Countries=",lme4::ngrps(int_homo)),
       x="",y="Redistributive preferences")
```

@@

ccoef <- list(homclass3_V_gc="Social distance (CWC)",
              "dclass3res_VIntermediate class (III+IV)"="Intermediate Class",
              "dclass3res_VWorking Class (V+VI+VII)"="Working Class",
              "gini_disp" = "Income inequality (Gini index)",
              loggdppercapita = "GDP/capita",
              wstate2="Size of the welfare state",
              "homclass3_V_gc:dclass3res_VIntermediate class (III+IV)" = "Distance*Intermediate Class",
              "homclass3_V_gc:dclass3res_VWorking Class (V+VI+VII)" = "Distance*Working Class",
              "homclass3_V_gc:gini_disp"= "Homogeneity*Income Inequality",
              "homclass3_V_gc:dclass3res_VIntermediate class (III+IV):gini_disp" = "Distance*Intermediate Class*Income Inequality",
              "homclass3_V_gc:dclass3res_VWorking Class (V+VI+VII):gini_disp" = "Distance*Working Class*Income Inequality")

@@

## Social Distance by ISEI

@@

dummy_df <- df1 %>% dplyr::select(id,
  egal = egal2,dclass3res_V,dclass3spo_V,homclass3_V_res,know_total,socialtrust,
  Q03pcm,edyears,female,agenum,religion,partner,union,workst,WEIGHT,region,
  "gini_disp"=wid_gini_disp,"gini_mkt",gv_spen,rel_red,abs_red,ilo_taxrev,
  ilo_govspe,d10d1=wid_p90p10,wid_p90p50,top10=wid_sharetop10,rgdpna,gdppercapita,
  oecd,country2, country, country3) %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
  dplyr::select(id)

@@

@@

# df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
dummy_df <- df1 %>% dplyr::select(id,
  egal = egal2,dclass3res_V,dclass3spo_V,homclass3_V_res,know_total,socialtrust,
  Q03pcm,edyears,female,agenum,religion,partner,union,workst,WEIGHT,region,
  "gini_disp"=wid_gini_disp,"gini_mkt",gv_spen,rel_red,abs_red,ilo_taxrev,
  ilo_govspe,d10d1=wid_p90p10,wid_p90p50,top10=wid_sharetop10,rgdpna,gdppercapita,
  oecd,country2, country, country3) %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
  dplyr::select(id)

df1$isei08sp <- car::recode(df1$isei08sp,"NA=0")
dfreg <- df1 %>% dplyr::select(id,
  egal = egal2,
  # starts_with("class3"),
  # starts_with("digclass3"),
  # starts_with("dclass3"),
  # starts_with("homclass"),
  dclass3res_V=isei08,
  dclass3spo_V=isei08sp,
  homclass3_V_res=socdist,
  know_total,
  socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  religion,
  partner,
  union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  "gini_mkt",
  gv_spen,
  rel_red,
  abs_red,
  ilo_taxrev,
  ilo_govspe,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  rgdpna,
  gdppercapita,
  oecd,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         gini_disp=scale(gini_disp),
         dclass3res_V=dclass3res_V/10,
         dclass3spo_V=dclass3spo_V/10,
         homclass3_V_res=homclass3_V_res/10)

@@

texreg(digclas1_macro,stars = c(0.001, 0.01, 0.05, 0.1),symbol = "+",scalebox = 0.60,
       caption = "Multilevel models for Income Inequality, Network segregation and Redistributive Preferences",
       custom.note = "Note: Models include sampling weights and individual level controls centered within cluster (group mean). Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,single.row = T,
       custom.coef.map = ccoef,
      # groups = list("Social Class (Ref.= Service Class)" = 2:3,"Macro-level factors"=4:6, "Distance*Social Class"=7:8, "Distance * Social Class * Income Inequality"=10:11),
       float.pos = "h!",
      custom.gof.rows = list("Controls"=rep("Yes",4)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups",
                           "Var: Group (Intercept)",
                           "Var: Group Soc. Distance",
                           "Var: Group ISEI",
                           "Cov: Group (Intercept), Soc. Distance",
                           "Cov: Group (Intercept), ISEI",                           
                           "Cov: Group Soc. Distance, ISEI",
                           "Var: Residual"),
       use.packages = F)
```

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | The escalating economic inequality and the current sanitary crisis have threatened social solidarity among citizens. However, these consequences have been unevenly experienced across the social structure where the upper and intermediate classes, benefiting from privileged access to social resources, have shown resilience. Conversely, working-class families have faced deteriorating material conditions, leading to a heightened sense of marginalization and increasing the demand for welfare support. Particularly, social resources through network ties have been suggested to have a direct link with an individual's welfare through instrumental and expressive outcomes, such as providing information or help in moments of need. Consequently, it has been argued that social isolation not only plays a role in terms of social resources, but higher levels of segregation can bolster opinions, suggesting that being isolated from other social classes can polarise attitudes mainly in the working and upper-middle classes. In this regard, based on the ISSP 2017 -  Social Networks (N=31,191),  multilevel estimations suggest that the influence of network homogeneity is conditional to social class, where the working and intermediate classes with high homogeneous networks hold higher redistributive preferences than the upper classes. Additionally, income inequality has been shown to decrease the polarizing effect of network segregation, mainly among the upper classes, when overall inequality is high. In sum, this study highlights the role of both social class and network segregation together to understand redistributive preferences, as well as their implications for policies and cohesion.
  | 
  | **Keywords**: network segregation, redistributive preferences, economic inequality
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

Individual data from the International Social Survey Programme (ISSP) 2017, titled "Social Networks and Social Resources," is utilized [@issp_networks_2017]. Established in 1984, the ISSP is a collaborative research project that conducts cross-national surveys on various topics related to social issues, attitudes, and values. Specifically, the ISSP 2017 survey includes questions on social ties and activities, evaluated through a position generator, as well as inquiries on social resources derived from network members. Additionally, respondents express their views on topics such as social trust and attitudes toward economic inequality. Initially, the dataset comprised 47,027 individual observations across 32 countries. However, Slovenia (SVN) was excluded from the analysis because the key dependent variable was not present in the questionnaire. Consequently, the present study comprises complete information on 31,191 individuals from 31 countries, representing a group of diverse societies in terms of socio-economic and institutional arrangements.

@@

\pagebreak 
# Additional analysis

## Income Inequality Ratios

@@

texreg(digclas1_macro,stars = c(0.001, 0.01, 0.05, 0.1),symbol = "+",scalebox = 0.50,
       caption = "Multilevel models for Income Inequality, Network segregation and Redistributive Preferences",
       custom.note = "Note: Models include sampling weights and individual level controls centered within cluster (group mean). Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,single.row = T,
       custom.coef.map = ccoef,
      groups = list("Social Class (Ref.= Service Class)" = 2:3,"Macro-level factors"=4:6, "Homogeneity*Social Class"=7:8, "Homogeneity * Social Class * Income Inequality"=10:11),
       float.pos = "h!",
      custom.gof.rows = list("Controls"=rep("Yes",4)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups",
                           "Var: Group (Intercept)",
                           "Var: Group Homogeneity",
                           "Var: Group Intermediate Class ",
                           "Var: Group Working Class",
                           "Cov: Group (Intercept), Homogeneity",
                           "Cov: Group (Intercept), Intermediate Class ",
                           "Cov: Group (Intercept), Working Class",
                           "Cov: Group Homogeneity, Intermediate Class",
                           "Cov: Group Homogeneity, Working Class",
                           "Cov: Group Intermediate Class, Working Class",
                           "Var: Residual"),
       use.packages = F)
```

@@

## Social distance

First, the average scores of the International Socio-Economic Index of Occupational Status (ISEI) [@ganzeboom_three_2003] for each occupation of the position generator are calculated. Second, the ISEI score of the respondent (R's) is subtracted from the average ISEI points of the personal network. For example, if the R's has an ISEI of 80 and the network ISEI is 50, the social distance will be 30 (80 - 50), a "upward" social distance. Another case could be 50 (R's) minus 80 (network), and the average social distance will be - 30 or "downward" social distance. In addition, When the distance is 0, the network is entirely homogeneous.

To facilitate the interpretation of the indicator, we have calculated a homogeneity indicator based on social distance:

1. The absolute values are calculated to represent the total distance to occupations respect to R's ISEI score.
2. Since we have 0 values representing absolute homogeneity, we rescale the variable by adding 1.
4. We have inverted the values to make higher values represent higher homogeneity.

Thus, higher values represent greater homogeneity regarding R's ISEI score in contrast to the average network ISEI score.

@@

### Social Distance by Social Class

@@

load(here::here("input/data/proc/study1_country.RData"));df1 <- df2
dfreg <- df1 %>% dplyr::select(
  egal = egal2,
  dclass3res_V,
  dclass3spo_V,
  homclass3_V_res=socdist,
  know_total,
  socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  partner,
  religion,
  union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  "gini_mkt",
  gv_spen,
  rel_red,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  rgdpna,
  gdppercapita,
  oecd,
  country2, country, country3,
) %>%
  mutate(logrgdpna = log(rgdpna),
         loggdppercapita=log(gdppercapita),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2,
         homclass3_V_res=homclass3_V_res/10) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN")
dfreg$dclass3spo_V <- car::recode(dfreg$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))

@@

# Step 1: Calculate absolute values
absolute_values <- abs(dfreg$homclass3_V_res)
# hist(absolute_values);summary(absolute_values)

# Step 2: Add 1 to all values to eliminate zeros
modified_values <- absolute_values + 0.1
# hist(modified_values);summary(modified_values)

# Step 3: Reverse the order of the scale
reversed_values <- max(modified_values) + 0.1 - modified_values
# hist(reversed_values);summary(reversed_values)

dfreg$homclass3_V_res <- reversed_values

@@

```{r fig1-marg-socdistclass, echo=FALSE, fig.cap='Average Marginal Effects of Social Distance conditionated by Social Class on Redistributive Preferences', fig.dim=c(10,5), warning=FALSE}
int_homo <- digclas1_VW_7
plot_slopes(digclas1_VW_7, variables = "homclass3_V_res", condition = list("dclass3res_V"))+
labs(caption = paste0("N=",nobs(int_homo),", Countries=",lme4::ngrps(int_homo)),
       x="Social Distance",y="Redistributive preferences")
```

@@

ccoef <- list(homclass3_V_gc="Social distance (CWC)",
              "dclass3res_VIntermediate class (III+IV)"="Intermediate Class",
              "dclass3res_VWorking Class (V+VI+VII)"="Working Class",
              "gini_disp" = "Income inequality (Gini index)",
              loggdppercapita = "GDP/capita",
              wstate2="Size of the welfare state",
              "homclass3_V_gc:dclass3res_VIntermediate class (III+IV)" = "Distance*Intermediate Class",
              "homclass3_V_gc:dclass3res_VWorking Class (V+VI+VII)" = "Distance*Working Class",
              "homclass3_V_gc:gini_disp"= "Distance*Income Inequality",
              "homclass3_V_gc:dclass3res_VIntermediate class (III+IV):gini_disp" = "Distance*Intermediate Class*Income Inequality",
              "homclass3_V_gc:dclass3res_VWorking Class (V+VI+VII):gini_disp" = "Distance*Working Class*Income Inequality")

@@

### Social Distance by ISEI

@@

dummy_df <- df1 %>% dplyr::select(id,
  egal = egal2,dclass3res_V,dclass3spo_V,homclass3_V_res,know_total,socialtrust,
  Q03pcm,edyears,female,agenum,religion,partner,union,workst,WEIGHT,region,
  "gini_disp"=wid_gini_disp,"gini_mkt",gv_spen,rel_red,abs_red,ilo_taxrev,
  ilo_govspe,d10d1=wid_p90p10,wid_p90p50,top10=wid_sharetop10,rgdpna,gdppercapita,
  oecd,country2, country, country3) %>%
  filter(country2 != "SVN") %>% 
  # filter(oecd == "OECD") %>%
  na.omit() %>% 
  dplyr::select(id)

@@

# Step 1: Calculate absolute values
absolute_values <- abs(dfreg$homclass3_V_res)
# hist(absolute_values);summary(absolute_values)

# Step 2: Add 1 to all values to eliminate zeros
modified_values <- absolute_values + 0.1
# hist(modified_values);summary(modified_values)

# Step 3: Reverse the order of the scale
reversed_values <- max(modified_values) + 0.1 - modified_values
# hist(reversed_values);summary(reversed_values) 
dfreg$homclass3_V_res <- reversed_values

@@

df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
dummy_df <- df1 %>% dplyr::select(id,
  egal = egal2,dclass3res_V,dclass3spo_V,homclass3_V_res,know_total,socialtrust,
  Q03pcm,edyears,female,agenum,religion,partner,union,workst,WEIGHT,region,
  "gini_disp"=wid_gini_disp,"gini_mkt",gv_spen,rel_red,abs_red,ilo_taxrev,
  ilo_govspe,d10d1=wid_p90p10,wid_p90p50,top10=wid_sharetop10,rgdpna,gdppercapita,
  oecd,country2, country, country3) %>%
  filter(country2 != "SVN") %>% 
  # filter(oecd == "OECD") %>%
  na.omit() %>% 
  dplyr::select(id)

df1$isei08sp <- car::recode(df1$isei08sp,"NA=0")
dfreg <- df1 %>% dplyr::select(id,
  egal = egal2,
  # starts_with("class3"),
  # starts_with("digclass3"),
  # starts_with("dclass3"),
  # starts_with("homclass"),
  dclass3res_V=isei08,
  dclass3spo_V=isei08sp,
  homclass3_V_res=socdist,
  know_total,
  socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  religion,
  partner,
  union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  "gini_mkt",
  gv_spen,
  rel_red,
  abs_red,
  ilo_taxrev,
  ilo_govspe,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  rgdpna,
  gdppercapita,
  oecd,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         gini_disp=scale(gini_disp),
         dclass3res_V=dclass3res_V/10,
         dclass3spo_V=dclass3spo_V/10,
         homclass3_V_res=homclass3_V_res/10) %>% 
  filter(id %in% dummy_df$id)

dfreg$wstate2 <- scale(scale(dfreg$abs_red*100) + scale(dfreg$ilo_taxrev) + scale(dfreg$ilo_govspe))
# Step 1: Calculate absolute values
absolute_values <- abs(dfreg$homclass3_V_res)
# hist(absolute_values);summary(absolute_values)

# Step 2: Add 1 to all values to eliminate zeros
modified_values <- absolute_values + 0.1
# hist(modified_values);summary(modified_values)

# Step 3: Reverse the order of the scale
reversed_values <- max(modified_values) + 0.1 - modified_values
# hist(reversed_values);summary(reversed_values)

dfreg$homclass3_V_res <- reversed_values

@@

texreg(digclas1_macro,stars = c(0.001, 0.01, 0.05, 0.1),symbol = "+",scalebox = 0.60,
       caption = "Multilevel models for Income Inequality, Social Distance and Redistributive Preferences",
       custom.note = "Note: Models include sampling weights and individual level controls centered within cluster (group mean). Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,single.row = T,
       custom.coef.map = ccoef,
      # groups = list("Social Class (Ref.= Service Class)" = 2:3,"Macro-level factors"=4:6, "Distance*Social Class"=7:8, "Distance * Social Class * Income Inequality"=10:11),
       float.pos = "h!",
      custom.gof.rows = list("Controls"=rep("Yes",4)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups",
                           "Var: Group (Intercept)",
                           "Var: Group Soc. Distance",
                           "Var: Group ISEI",
                           "Cov: Group (Intercept), Soc. Distance",
                           "Cov: Group (Intercept), ISEI",                           
                           "Cov: Group Soc. Distance, ISEI",
                           "Var: Residual"),
       use.packages = F)
```
